---
title: "Exploratory Data Analysis (Course Project 1)"
output:
  html_document:
    df_print: paged
---

The goal of this course project is to examine how household energy usage varies over a 2-day period in February, 2007. For this, we need to reconstruct four exploratory plots, all of which were constructed using the base plotting system.


This project includes:

1. R scripts downloading the data and making the plots (download.R, plot1.R, plot2.R, plot3.R, plot4.R)
2. code book describing the data and any transformations performed to make the plots (CodeBook.Rmd)
3. Four plots (plot1.png, plot2.png, plot3.png and plot4.png)
4. README file explaining how all of the scripts work (README.md)

All the analyses were performed using R version 4.0.2 (2020-06-22) and RStudio version 1.2.5033.

# Description of the data

For this project, we use the "Individual household electric power consumption Data Set" downloaded from the [UC Irvine Machine Learning Repository](http://archive.ics.uci.edu/ml/index.php), a popular repository for machine learning datasets. It represents the measurements of electric power consumption in one household with a one-minute sampling rate over a period of almost 4 years. Different electrical quantities and some sub-metering values are available, but in this project we focus on these nine variables:

1. Date: Date in format dd/mm/yyyy
2. Time: time in format hh:mm:ss
3. Global_active_power: household global minute-averaged active power (in kilowatt)
4. Global_reactive_power: household global minute-averaged reactive power (in kilowatt)
5. Voltage: minute-averaged voltage (in volt)
6. Global_intensity: household global minute-averaged current intensity (in ampere)
7. Sub_metering_1: energy sub-metering No. 1 (in watt-hour of active energy). It corresponds to the
kitchen, containing mainly a dishwasher, an oven and a microwave (hot plates are not electric but gas powered).
8. Sub_metering_2: energy sub-metering No. 2 (in watt-hour of active energy). It corresponds to the laundry room, containing a washing-machine, a tumble-drier, a refrigerator and a light.
9. Sub_metering_3: energy sub-metering No. 3 (in watt-hour of active energy). It corresponds to an electric water-heater and an air-conditioner.

## Downloading the data

First, create and clone github repository for your project. Then, set/get the working directory to your cloned github repo on your computer. Use the following lines of code to download the dataset and unzip it as "household_power_consumption.txt":

```{r downloading, echo=FALSE}
fileURL <- "https://d396qusza40orc.cloudfront.net/exdata%2Fdata%2Fhousehold_power_consumption.zip"
temp <- tempfile(); download.file(fileURL,temp)
file.copy(from = temp, to = paste0(getwd(),"/raw_data.zip")) # download zip file as raw_data.zip
unlink(temp) # delete the temp file or use file.remove(temp)
unzip("raw_data.zip") # unzip the raw_data.zip as "household_power_consumption.txt"
```

## Loading the data

```{r loading, echo=FALSE}
# The dataset has 2,075,259 rows and 9 variables
data <- read.table("household_power_consumption.txt", sep=";", header=TRUE, na.strings=c("NA", " ", "?"))
# Convert Date to Date class
data$Date <- format(as.Date(data$Date, "%d/%m/%Y"))
# Convert Time to Time class
#library(lubridate); data$Time <- hms(data$Time)
data$Time <- format(strptime(data$Time, "%H:%M:%S"),"%H:%M:%S")
```

## Subsetting data for dates dates 2007-02-01 and 2007-02-02

```{r subbsetting, echo=FALSE}
subdata <- subset(data, data$Date == "2007-02-01" | data$Date == "2007-02-02")
subdata$DateTime <- as.POSIXct(paste(subdata$Date, subdata$Time))
active_power <- as.numeric(subdata$Global_active_power)
```

## Making plots

This chunk of code makes plot1.png (see plot1.R)

```{r plot1, echo=FALSE}
png(filename = "plot1.png", width = 480, height = 480, units = "px")
par(bg=NA)
hist(subdata$Global_active_power, col="red", xlab="Global Active Power (killowatts)",
     main="Global Active Power")
dev.off()
```

This chunk of code makes plot2.png (see plot2.R)

```{r plot2, echo=FALSE}
png(filename = "plot2.png", width = 480, height = 480, units = "px")
par(bg=NA)
plot(subdata$DateTime, active_power, type="l", xlab="", ylab="Global Active Power (kilowatts)")
dev.off()
```

This chunk of code makes plot3.png (see plot3.R)

```{r plot3, echo=FALSE}
png(filename = "plot3.png", width = 480, height = 480, units = "px")
par(bg=NA)
plot(subdata$DateTime, subdata$Sub_metering_1, type = "l", xlab = "", ylab = "Energy sub metering")
lines(subdata$DateTime, subdata$Sub_metering_2, type = "l", col = "red")
lines(subdata$DateTime, subdata$Sub_metering_3, type = "l", col = "blue")
legend("topright", c("Sub_metering_1", "Sub_metering_2", "Sub_metering_3"), 
       lwd = 2.5, col = c("black", "red", "blue"))
dev.off()

# Or use points
png(filename = "plot3.png", width = 480, height = 480, units = "px")
par(bg=NA)
plot(subdata$DateTime, subdata$Sub_metering_1, type = "l", xlab = "", ylab = "Energy sub metering")
points(subdata$DateTime, subdata$Sub_metering_2, type = "l", col = "red")
points(subdata$DateTime, subdata$Sub_metering_3, type = "l", col = "blue")
legend("topright", c("Sub_metering_1", "Sub_metering_2", "Sub_metering_3"), 
       lwd = 2.5, col = c("black", "red", "blue"))
dev.off()
```

This chunk of code makes plot4.png (see plot4.R)

```{r plot4, echo=FALSE}
png(filename = "plot4.png", width = 480, height = 480, units = "px")
par(mfrow = c(2,2), mar = c(4,4,2,1), bg=NA)
plot(subdata$DateTime, subdata$Global_active_power, type = "l",
     xlab = "", ylab = "Global Active Power")
plot(subdata$DateTime, subdata$Voltage, type = "l", xlab = "datetime", ylab = "Voltage")
plot(subdata$DateTime, subdata$Sub_metering_1, type = "l", xlab = "", ylab = "Energy sub metering")
lines(subdata$DateTime, subdata$Sub_metering_2, type = "l", col = "red")
lines(subdata$DateTime, subdata$Sub_metering_3, type = "l", col = "blue")
legend("topright", c("Sub_metering_1", "Sub_metering_2", "Sub_metering_3"), 
       lwd = 2.5, col = c("black", "red", "blue"))
plot(subdata$DateTime, subdata$Global_reactive_power, type = "l", 
     xlab = "datetime", ylab = "Global_reactive_power")
dev.off()
```
